Aldebaran~b's temperate past uncovered in planet search data

### 2018-02-05 paper/aldebaran.tex
title
planet not star
Fixed typos

\title{Detection of Oscillations in Aldebaran with Ground-Based Observations}

@@

%As can be seen from the scaling relations, asteroseismology can provide constraint on Mass, Radius, and surface gravity if \Dnu, \numax, and \teff can be measured.
The ability to measure \numax and \Dnu depends on the length and sampling rate of a data set.
In practise, for the higher luminosity red giants it is more straight forward to measure \numax than \Dnu.  Typical values for \numax\ range from $0.1-20 \rm\,\mu Hz$ for luminous giants and $20-50 \rm\,\mu Hz$ for stars near the red clump.  For the evolutionary stages, \Dnu typically ranges from $0.02-3 \rm\,\mu Hz$ and $3-7 \rm\,\mu Hz$ depending on stellar mass and \teff
%(or metallicity which is considered more fundamental)
\citep[e.g.,][]{2011A&A...525L...9M, 2013A&A...559A.137M}.  Hence, \numax requires less frequency resolution than \Dnu to establish a good measurement.

@@

Radial velocity variations intrinsic to the star, such as those caused by stellar oscillations, have previously limited the precision with which red giant planets have been studied \citep{2005PASJ...57...97S}. We aim to model out these noise processes and use them recover asteroseismic information, as well as to improve our estimates of the orbital parameters of the planet. It is easy to observe Aldebaran and similarly-bright stars with ground-based spectroscopic instruments, typically only requiring short exposures that can be obtained even under adverse observing conditions. There is indeed a considerable archive of such observations already, as a legacy of radial velocity (RV) surveys conducted to find exoplanets. In most cases, however, these have not so far been useful for asteroseismology, because these RV data are sparsely and irregularly sampled. Because we have to pause observations during the day, during poor weather conditions, or simply when targets of higher priority are being observed, we get time series which may have significant and uneven gaps. This introduces a window-function effect: the power spectrum as constructed for example by a Fourier transform, or a Lomb-Scargle Periodogram \citep{lomb,scargle}, is convolved with the Fourier transform of the window function, introducing strong sidelobes adjacent to real frequency peaks and causing crosstalk between adjacent frequency channels. This imposes significant limitations both on the signal-to-noise and frequency resolution of power spectra derived from linear methods such as the Lomb-Scargle periodogram, and in practice makes asteroseismology with these conventional approaches difficult or impossible from the ground for stars with oscillation periods ranging from $\sim 12$~h to $\sim$~a few days.

If we apply nonlinear statistical inference methods this situation can be
improved. \citet{brewer2009} show that a system of driven, damped harmonic
oscillators such as we encounter in asteroseismology can be statistically
modelled as a Gaussian Process, with a covariance kernel consisting of a sum of
damped sinusoids. The hyperparameters of this process encode features of the
power spectral density distributions and are insensitive to the window function. Gaussian processes with quasi-periodic kernels have previously been used to re-analyse RV data for main sequence stars \citep{2014MNRAS.443.2517H,2015MNRAS.452.2269R}, where the noise is from stellar activity, but have not previously been applied to red giant asteroseismology.
Unfortunately, these methods are impractical for long time series, as the computational
cost of evaluating the standard Gaussian Process likelihood function scales as
$\mathcal{O}(n^3)$.

Fortunately, the damped and driven harmonic oscillator GP can be written as the
solution of a class of stochastic ordinary differential equations, the
Continuous Auto-Regressive Moving Average (CARMA) models, whose likelihoods can
be evaluated in linear time. For problems such as this, we therefore have access
to these powerful computational tools for inferring their power spectral
densities.  Our treatment of CARMA models is described more fully in Appendix
\ref{carmatheory}.   \citet{Kelly2014} also gives more details on CARMA
processes in an astronomical context, and shows how a state-space model of the
process $y(t)$ can be tracked through a time-series of uncertain observations
$y_k = y\left( t_k \right) + \epsilon_k$, with Gaussian noise $\epsilon_k$ of
zero mean and known (heteroskedastic) variance, using a Kalman filter to produce
a likelihood function $p\left( y_k \mid \sigma, r_i, b_j \right)$ depending on
the amplitude of the stochastic forcing, $\sigma$, the (possibly complex)
eigenfrequencies of the ODE, $r_i$, and parameters describing the power spectrum
of the stochastic forcing, $b_j$ in $\mathcal{O}\left( n \right)$ computational
time.  Additionally, \citet{Foreman-Mackey2017} show how a dimensional expansion
can be used to reduce the standard Gaussian Process covariance matrix for this
same process,  with autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under
observations at times $t_k$, to a banded form that can then be
Cholesky-decomposed in the standard Gaussian Process likelihood function in
$\mathcal{O}\left( n \right)$ time.

@@

From this pilot study with Aldebaran, have shown that with limited quantities of data of limited quality, we can nevertheless do asteroseismology from the ground and obtain very precise estimates of stellar parameters. Furthermore, improvements in the algorithm may also permit the detection of not only~\numax but also~\Dnu from similar measurements, permitting more sophisticated asteroseismic analysis. Sufficient radial velocity data either already exist, or are trivial to obtain, in order to do this for essentially all bright giants; while for solar-like stars, these new methods will allow significantly relaxed observing requirements. One rich archive is the series of observations with the Hamilton \'{E}chelle Spectrograph at Lick Observatory, studying 373 G~and K~giants \citep[e.g.][]{frink2001,hekker2006,ortiz2016}. The method may also be useful in detecting rotational or other periodic variability in the LCES HIRES/Keck Precision Radial Velocity Exoplanet Survey \citep{butler17} of 1,624 FGKM dwarfs, although the time sampling is by design too coarse to detect $p$-mode oscillations in these stars. Furthermore, CARMA models and related methods will enhance deep, all-sky, sparse photometric surveys: an immediate future test of this will be from \hipparcos; while only~58 epochs of photometry are available for Aldebaran, at a sampling we find to be insufficient for our purposes, stars at higher latitudes may often have 150--200 epochs and more even sampling \citep{hipparcos_phot}, and the~14 bright~K giants in \hipparcos noted by \citet{bedding2000} are an ideal first test case. This can naturally be extended to \emph{Gaia} in space \citep{gaia}, or \emph{LSST} from the ground \citep{dmt,lsst,lsstbook}, from which many thousands of new asteroseismic determinations will be possible. In future RV surveys, it may be possible to beat the \citet{2005PASJ...57...97S} $\sim 30$~m/s RV precision limit for red giants by taking many closely spaced observations and modelling-out the effects of stellar oscillations: this will allow us to dig deeper into this intrinsic stellar noise to detect less-massive planets around these stars.

@@

The \kepler detector saturates for stars brighter than the $\sim11^\text{th}$ magnitude. Nevertheless, the excess flux deposited in a saturated pixel spills conservatively up and down the pixel column, such that it is possible to sum this `bleed column' for bright stars and still obtain precise photometry, such as was done for the brightest star in the nominal \kepler mission, $\theta$~Cyg \citep[$V = 4.48$;][]{guzik2011,thetacygwhite,guzik2016}. There are two main reasons why this is not possible for all bright stars in general. First, because the on-board data storage and downlink bandwidth from \kepler are limited, it is often not desirable to download the large number of pixels that are required for such bright stars. Second, if the bleed column for a sufficiently bright star reaches the edge of the chip, flux spills over and is not conserved, imposing a hard brightness limit that depends on the distance to the detector edge.
Collateral `smear' data, which are collected to help calibrate the photometric bias from stars sharing the same column as a target, can be used to reconstruct light curves for un-downloaded bright stars and thereby avoid bandwidth constraints \citep{k2smear}, but these data are still rendered unusable if the bleed column falls off the edge of the chip and contaminates the smear rows.

@@

As a final step to reduce residual uncorrected systematics, we apply the \textsc{k2sc} \citep{k2sc} Gaussian Process-based systematics correction code to the initial halo light curve, but the effect of this in the present case for Aldebaran is minimal.
There is somewhat higher than usual residual noise at harmonics of $4 c/d$ (the satellite thruster firing frequency), but this is nevertheless very small in comparison to the signal from Aldebaran, and may be ascribed to the large fraction of the pixel mask occupied by the bleed column from this extremely bright star.

@@

To ensure the values are self-consistent, when a literature value was chosen for temperature, we took the stellar metallicity from the same source i.e. matched pairs of temperature and metallicity. The final constraint is the stellar luminosity, which may be estimated as follows (e.g. see \citealt{pijpers2003}):
\begin{multline}
\log_{10} \frac{L}{L_{\odot}} = 4.0+
0.4 M_{{\textrm{bol}},\odot} -2.0 \log_{10} {\pi [{\textrm{mas}]}} \\-0.4(V-A_V + BC(V)).
\label{eqn:lum}
\end{multline}
The solar bolometric magnitude $M_{\textrm{bol},\odot}=4.73$ is taken from \cite{Torres2010}, from which we also take the polynomial expression for the bolometric correction $BC(V)$. We assume extinction $A_V$ to be zero.

>>>>

\title{Aldebaran~b's temperate past uncovered in planet search data}

@@

%As can be seen from the scaling relations, asteroseismology can provide constraint on Mass, Radius, and surface gravity if \Dnu, \numax, and \teff can be measured.
The ability to measure \numax and \Dnu depends on the length and sampling rate of a data set.
In practice, for the higher luminosity red giants it is more straightforward to measure \numax than \Dnu.  Typical values for \numax\ range from $0.1-20 \rm\,\mu Hz$ for luminous giants and $20-50 \rm\,\mu Hz$ for stars near the red clump.  For the evolutionary stages, \Dnu typically ranges from $0.02-3 \rm\,\mu Hz$ and $3-7 \rm\,\mu Hz$ depending on stellar mass and \teff
%(or metallicity which is considered more fundamental)
\citep[e.g.,][]{2011A&A...525L...9M, 2013A&A...559A.137M}.  Hence, \numax requires less frequency resolution than \Dnu to establish a good measurement.

@@

Radial velocity variations intrinsic to the star, such as those caused by stellar oscillations, have previously limited the precision with which red giant planets have been studied \citep{2005PASJ...57...97S}. We aim to model out these noise processes and use them to recover asteroseismic information, as well as to improve our estimates of the orbital parameters of the planet. It is easy to observe Aldebaran and similarly-bright stars with ground-based spectroscopic instruments, typically only requiring short exposures that can be obtained even under adverse observing conditions. There is indeed a considerable archive of such observations already, as a legacy of radial velocity (RV) surveys conducted to find exoplanets. In most cases, however, these have not so far been useful for asteroseismology because these RV data are sparsely and irregularly sampled. Because we have to pause observations during the day, during poor weather conditions, or simply when targets of higher priority are being observed, we get time series which may have significant and uneven gaps. This introduces a window-function effect: the power spectrum as constructed for example by a Fourier transform, or a Lomb-Scargle Periodogram \citep{lomb,scargle}, is convolved with the Fourier transform of the window function, introducing strong sidelobes adjacent to real frequency peaks and causing crosstalk between adjacent frequency channels. This imposes significant limitations both on the signal-to-noise and frequency resolution of power spectra derived from linear methods such as the Lomb-Scargle periodogram, and in practice makes asteroseismology with these conventional approaches difficult or impossible from the ground for stars with oscillation periods ranging from $\sim 12$~h to $\sim$~a few days.

If we apply nonlinear statistical inference methods this situation can be
improved. \citet{brewer2009} show that a system of driven, damped harmonic
oscillators such as we encounter in asteroseismology can be statistically
modelled as a Gaussian Process (GP), with a covariance kernel consisting of a sum of
damped sinusoids. The hyperparameters of this process encode features of the
power spectral density distributions and are insensitive to the window function. Gaussian processes with quasi-periodic kernels have previously been used to re-analyse RV data for main sequence stars \citep{2014MNRAS.443.2517H,2015MNRAS.452.2269R}, where the noise is from stellar activity, but have not previously been applied to red giant asteroseismology.
Unfortunately, these methods are impractical for long time series, as the computational
cost of evaluating the standard Gaussian Process likelihood function scales as
$\mathcal{O}(n^3)$.

Fortunately, the damped and driven harmonic oscillator GP can be written as the
solution of a class of stochastic ordinary differential equations, the
Continuous Auto-Regressive Moving Average (CARMA) models, whose likelihoods can
be evaluated in linear time. For problems such as this, we therefore have access
to these powerful computational tools for inferring their power spectral
densities.  Our treatment of CARMA models is described more fully in Appendix
\ref{carmatheory}.   \citet{Kelly2014} also give more details on CARMA
processes in an astronomical context, and show how a state-space model of the
process $y(t)$ can be tracked through a time-series of uncertain observations
$y_k = y\left( t_k \right) + \epsilon_k$, with Gaussian noise $\epsilon_k$ of
zero mean and known (heteroskedastic) variance, using a Kalman filter to produce
a likelihood function $p\left( y_k \mid \sigma, r_i, b_j \right)$ depending on
the amplitude of the stochastic forcing, $\sigma$, the (possibly complex)
eigenfrequencies of the ODE, $r_i$, and parameters describing the power spectrum
of the stochastic forcing, $b_j$ in $\mathcal{O}\left( n \right)$ computational
time.  Additionally, \citet{Foreman-Mackey2017} show how a dimensional expansion
can be used to reduce the standard Gaussian Process covariance matrix for this
same process,  with autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under
observations at times $t_k$, to a banded form that can then be
Cholesky-decomposed in the standard Gaussian Process likelihood function in
$\mathcal{O}\left( n \right)$ time.

@@

From this pilot study with Aldebaran, we have shown that with limited quantities of data of limited quality, we can nevertheless do asteroseismology from the ground and obtain very precise estimates of stellar parameters. Furthermore, improvements in the algorithm may also permit the detection of not only~\numax but also~\Dnu from similar measurements, permitting more sophisticated asteroseismic analysis. Sufficient radial velocity data either already exist, or are trivial to obtain, in order to do this for essentially all bright giants; while for solar-like stars, these new methods will allow significantly relaxed observing requirements. One rich archive is the series of observations with the Hamilton \'{E}chelle Spectrograph at Lick Observatory, studying 373 G~and K~giants \citep[e.g.][]{frink2001,hekker2006,ortiz2016}. The method may also be useful in detecting rotational or other periodic variability in the LCES HIRES/Keck Precision Radial Velocity Exoplanet Survey \citep{butler17} of 1,624 FGKM dwarfs, although the time sampling is by design too coarse to detect $p$-mode oscillations in these stars. Furthermore, CARMA models and related methods will enhance deep, all-sky, sparse photometric surveys: an immediate future test of this will be from \hipparcos; while only~58 epochs of photometry are available for Aldebaran, at a sampling we find to be insufficient for our purposes, stars at higher latitudes may often have 150--200 epochs and more even sampling \citep{hipparcos_phot}, and the~14 bright~K giants in \hipparcos noted by \citet{bedding2000} are an ideal first test case. This can naturally be extended to \emph{Gaia} in space \citep{gaia}, or \emph{LSST} from the ground \citep{dmt,lsst,lsstbook}, from which many thousands of new asteroseismic determinations will be possible. In future RV surveys, it may be possible to beat the \citet{2005PASJ...57...97S} $\sim 30$~m/s RV precision limit for red giants by taking many closely spaced observations and modelling-out the effects of stellar oscillations: this will allow us to dig deeper into this intrinsic stellar noise to detect less-massive planets around these stars.

@@

The \kepler detector saturates for stars brighter than the $\sim11^\text{th}$ magnitude. Nevertheless, the excess flux deposited in a saturated pixel spills conservatively up and down the pixel column, such that it is possible to sum this `bleed column' for bright stars and still obtain precise photometry, such as was done for the brightest star in the nominal \kepler mission, $\theta$~Cyg \citep[$V = 4.48$;][]{guzik2011,thetacygwhite,guzik2016}. There are two main reasons why this is not possible for all bright stars in general. First, because the on-board data storage and downlink bandwidth from \kepler are limited, it is often not desirable to store and download the large number of pixels that are required for such bright stars. Second, if the bleed column for a sufficiently bright star reaches the edge of the chip, flux spills over and is not conserved, imposing a hard brightness limit that depends on the distance to the detector edge.
Collateral `smear' data, which are collected to help calibrate the photometric bias from stars sharing the same column as a target, can be used to reconstruct light curves for un-downloaded bright stars and thereby avoid bandwidth constraints \citep{k2smear}, but these data are still rendered unusable if the bleed column falls off the edge of the chip and contaminates the smear rows.

@@

As a final step to reduce residual uncorrected systematics, we apply the \textsc{k2sc} \citep{k2sc} Gaussian Process-based systematics correction code to the initial halo light curve, but the effect of this in the present case for Aldebaran is minimal.
There is somewhat higher than usual residual noise at harmonics of $46\,\mu\mathrm{Hz}$ ($4\,\mathrm{d}^{-1}$, the satellite thruster firing frequency), but this is nevertheless very small in comparison to the signal from Aldebaran, and may be ascribed to the large fraction of the pixel mask occupied by the bleed column from this extremely bright star.

@@

To ensure the values are self-consistent, when a literature value was chosen for temperature, we took the stellar metallicity from the same source i.e. matched pairs of temperature and metallicity. The final constraint is the stellar luminosity, which may be estimated as follows (e.g. see \citealt{pijpers2003}):
\begin{multline}
\log_{10} \frac{L}{L_{\odot}} = 4.0+
0.4 M_{{\textrm{bol}},\odot} -2.0 \log_{10} {\pi [{\textrm{mas}]}} -0.4(V-A_V + BC(V)).
\label{eqn:lum}
\end{multline}
The solar bolometric magnitude $M_{\textrm{bol},\odot}=4.73$ is taken from \cite{Torres2010}, from which we also take the polynomial expression for the bolometric correction $BC(V)$. We assume extinction $A_V$ to be zero.
